% chapter included in vclmanual.tex
\documentclass[vcl_manual.tex]{subfiles}
\begin{document}


\section{File list}
%\label{FileList}
\flushleft

\begin{longtable}[l]{|p{40mm}|p{100mm}|}
\endfirsthead
\label{table:fileList} \\
\endhead
\hline
\bfseries File name & \bfseries Purpose \\ \hline
manual/vcl\_manual.pdf & Instruction manual (this file) \\ \hline

vectorclass.h & Top-level C++ header file. This will include several other header files, according to the indicated instruction set \\ \hline

instrset.h & Detection of which instruction set the code is compiled for, 
and functions that depend on the instruction set. This file also contains various common definitions and templates. Included by vectorclass.h \\ \hline

vectori128.h & Defines classes, operators and functions for integer vectors with a total size of 128 bits. Included by vectorclass.h \\ \hline

vectori256.h & Defines classes, operators and functions for integer vectors with a total size of 256 bits for the AVX2 instruction set. Included by vectorclass.h if appropriate \\ \hline

vectori256e.h & Defines classes, operators and functions for integer vectors with a total size of 256 bits for instruction sets lower than AVX2. Included by vectorclass.h if appropriate \\ \hline

vectori512.h & Defines classes, operators and functions for vectors of 32-bit and 64-bit integers with a total size of 512 bits for the AVX512F instruction set. Included by vectorclass.h if appropriate \\ \hline

vectori512e.h & Defines classes, operators and functions for vectors of 32-bit and 64-bit integers with a total size of 512 bits for instruction sets lower than AVX512F. Included by vectorclass.h if appropriate \\ \hline

vectori512s.h & Defines classes, operators and functions for vectors of 8-bit and 16-bit integers with a total size of 512 bits for the AVX512BW instruction set. Included by vectorclass.h if appropriate \\ \hline

vectori512se.h & Defines classes, operators and functions for vectors of 8-bit and 16-bit integers  with a total size of 512 bits for instruction sets lower than
AVX512BW. Included by vectorclass.h if appropriate \\ \hline

vectorf128.h & Defines classes, operators and functions for floating point vectors with a total size of 128 bits. Included by vectorclass.h \\ \hline

vectorf256.h & Defines classes, operators and functions for floating point vectors with a total size of 256 bits for the AVX and later instruction sets. Included by vectorclass.h if appropriate \\ \hline

vectorf256e.h & Defines classes, operators and functions for floating point vectors with a total size of 256 bits for instruction sets lower than AVX. Included by vectorclass.h if appropriate \\ \hline

vectorf512.h & Defines classes, operators and functions for floating point vectors with a total size of 512 bits for the AVX512F and later instruction sets. Included by vectorclass.h if appropriate \\ \hline

vectorf512e.h & Defines classes, operators and functions for floating point vectors with a total size of 512 bits for instruction sets lower than AVX512F. Included by vectorclass.h if appropriate \\ \hline

vectorfp16.h & Defines classes, operators and functions for half precision floating point vectors of all sizes, including mathematical functions, for AVX512-FP16 \\ \hline

vectorfp16e.h & Defines emulating classes, operators and functions for half precision floating point vectors of all sizes, including mathematical functions, for processors without AVX512-FP16 \\ \hline

vector\_convert.h & Defines functions for conversion between different vector sizes, as well as some generic function templates. \\ \hline

vectormath\_exp.h & Optional inline mathematical functions: power, logarithms and exponential functions \\ \hline

vectormath\_trig.h & Optional inline mathematical functions: trigonometric and inverse trigonometric functions \\ \hline

vectormath\_hyp.h & Optional inline mathematical functions: hyperbolic and inverse hyperbolic functions \\ \hline

vectormath\_common.h & Common definitions for vectormath\_exp.h, vectormath\_trig.h and vectormath\_hyp.h \\ \hline

vectormath\_lib.h & Optional header file for external mathematical vector function library \\ \hline

instrset\_detect.cpp & Optional functions for detecting which instruction set is supported at runtime \\ \hline

dispatch\_example.cpp & Example of how to make automatic CPU dispatching \\ \hline

LICENSE & Apache 2.0 license \\ \hline

changelog.txt & VCL version history \\ \hline

miscellaneous/svmlpatch & Folder containing the library svmlpatch.lib as well as the source code to build it. Used for fixing a compatibillity issue with Intel SVML library in 64-bit Windows \\ \hline

testbench & Folder containing test bench files for testing the VCL library. This is used in the development of VCL, and is not needed by programmers using the VCL. Includes code and documentation. \\ \hline

\end{longtable}
%\end{tabular}
\vspacesmall



\end{document}
