% chapter included in vclmanual.tex
\documentclass[vcl_manual.tex]{subfiles}
\begin{document}

\chapter{Add-on packages}\label{chap:AddOnPackages}
\flushleft

Various extra packages are available with code for special applications. 
These packages are stored at 
\url{https://github.com/vectorclass/add-on}.
Manuals are included with each package. The add-on packages for VCL include:

\begin{description}

\item[Container classes.] 
   Container class templates for storing arrays of vectors. More efficient than the standard C++ container class templates. \newline
   This package also contains a class template for matrices where matrix rows are stored as VCL vectors. Various functions are included for accessing matrix elements and rows and for packing and unpacking matrix data.

\item[Random number generator.] 
   A high-quality pseudo random number generator. Capable of generating random integer and floating point vectors. Suitable for large multi-threaded applications.

\item[Decimal string conversion.] 
   Converts integer vectors to and from comma-separated lists in human-readable decimal ASCII form. Useful for reading and writing comma-separated files.

\item[3-dimensional vectors.] 
   Defines 3-dimensional vectors for use in geometry and physics. 
   Includes operators and functions for addition, multiplication, dot product, cross product, and rotation.
   
\item[Complex number vectors.]    
   Defines complex number vectors for use in mathematics and electronics. 
   Includes operators for add, subtract, multiply, divide, and conjugate, as well as functions such as complex square root, exponential function, and logarithm.
   
\item[Quaternions.]    
   Defines quaternions (hypercomplex numbers) for use in mathematics.
   Includes operators for add, subtract, multiply, divide, conjugate, etc.


% add more packages here


\end{description}
\vspacesmall


\end{document}